---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
```{r}
library(ggplot2)
#read data from the zipped file
data <- read.csv(unz("activity.zip", "activity.csv"))
#drop rows with NA values
data.no.na <- data[!is.na(data$steps), ]
```


## What is mean total number of steps taken per day?
```{r}
#split data for each day
data.splitted.date <- split(data.no.na$steps, factor(data.no.na$date))
#find total number of steps for each day
total.per.day <- sapply(data.splitted.date, sum)
#caluclate mean and median number of steps per day
mean.steps.per.day <- mean(total.per.day)
median.steps.per.day <- median(total.per.day)
```
Mean number of steps per day is equal to $`r mean.steps.per.day`$.
Median number of steps per day is equal to $`r median.steps.per.day`.$

```{r}
#Plot histogram of the total number of steps
qplot(total.per.day, geom="histogram", binwidth=2000, xlab="Number of steps per day")
```


## What is the average daily activity pattern?
```{r}
#split data for time intervals
data.splitted.interval <- split(data.no.na$steps, factor(data.no.na$interval))
#caluclate mean and meadian number of steps per time interval across different days
mean.steps.per.interval <- sapply(data.splitted.interval, mean)
median.steps.per.interval <- sapply(data.splitted.interval, median)

#time interval with maximum average number if steps
index.max.steps <- as.numeric(which(mean.steps.per.interval == max(mean.steps.per.interval)))

max.interval <- unique(data$interval)[index.max.steps]

ggplot() + 
  geom_line(aes(unique(data$interval), mean.steps.per.interval)) + 
  xlab("Time interval") + 
  ylab("Mean of number of steps per interval") +
  geom_vline(xintercept=unique(data.no.na$interval)[index.max.steps], col="red")
```

Interval with maximum number of steps starts at $`r max.interval`$ and continues for next $5$ seconds.

## Imputing missing values

NA values filled with mean over the time intervals averaged over rest of the days.

```{r}
#number of rows with missing values:
rows.with.na <- sum(complete.cases(data) == FALSE)

#replace missed values with mean value per corresponding interval over all days
data.filled <- data
for (i in 1:nrow(data)) {  
  row <- data[i, ]      
  if (is.na(row["steps"])) {    
    index.rows <- which(unique(data.no.na$interval)==as.numeric(row["interval"]))
    new.steps.value <- as.numeric(mean.steps.per.interval[index.rows])
    data.filled$steps[i] <- new.steps.value    
    }
  }

#split data for each day
data.filled.splitted.date <- split(data.filled$steps, factor(data.filled$date))
#find total number of steps for each day
total.filled.per.day <- sapply(data.filled.splitted.date, sum)
#caluclate mean and median number of steps per day
mean.filled.steps.per.day <- mean(total.filled.per.day)
median.filled.steps.per.day <- median(total.filled.per.day)
```
Mean number of steps per day is equal to $`r mean.steps.per.day`$.
Median number of steps per day is equal to $`r median.steps.per.day`.$

```{r}
#Plot histogram of the total number of steps
library(gridExtra)
plot1 <- qplot(total.filled.per.day, geom="histogram", binwidth=2000, xlab="Number of steps per day, where NA filled")
plot2 <- qplot(total.per.day, geom="histogram", binwidth=2000, xlab="Number of steps per day, where NA dropped")
grid.arrange(plot1, plot2, nrow=2)
```

Total number of rows with NA is $`r rows.with.na`$. After filling na values for the averages for 
corresponding period we get total mean value `r mean.filled.steps.per.day`, median `r median.filled.steps.per.day`. As we can see, mean, median and histragram for the total of number of steps per day are almost identical to the unfilled values.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
#add column to the dataframe that corresponds to the day of the week
data.filled["weekday"] <- weekdays(as.Date(data.filled$date))
#add column to the dataframe that will tell if it is weekend ot not
data.filled["weekend"] <- data.filled$weekday %in% c("Saturday", "Sunday")

#split data into weekend and weekdays variables
weekend <- data.filled[data.filled$weekend, ]
weekday <- data.filled[!data.filled$weekend, ]

#split data for time intervals
mean.weekend <- sapply(split(weekend$steps, factor(weekend$interval)), mean)
mean.weekday <- sapply(split(weekday$steps, factor(weekday$interval)), mean)

plot1 <- qplot(unique(weekend$interval), 
               mean.weekend, 
               xlab="Time interval", 
               ylab=ylab("Mean of number of steps"), 
               geom="line",
               main="Weekend")

plot2 <- qplot(unique(weekday$interval), 
               mean.weekday, 
               xlab="Time interval", 
               ylab=ylab("Mean of number of steps"), 
               geom="line",
               main="Weekday")
grid.arrange(plot1, plot2, nrow=2)
```

Activities on the weekday and weekend look different.